One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-07-27 papers/1_dmatpol/paper-r3.tex
Field Gradient removed fromthe paper

In this work, the electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibilities
interacting with electric fields. The analytical approximate expressions were derived
at the Hartree\hyp{}Fock theory, which served as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron density of water molecule in spatially non\hyp{}uniform electric field distribution
generetated by point charges. The electric field\hyp{}induced interaction energies,
dipole moments and quadrupole moments
are quantitatively described by the present theory in weak and moderate electric fields (0.003 -- 0.02 a.u.).
%It is believed that the possibility of extension of the proposed model with high\hyp{}order susceptibilities
%opens new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems.
It is believed that the proposed model, after extensive development on its optimization, could open
new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems.
\end{abstract}

@@

Molecular\hyp{}level exploration of condensed phase phenomena becomes increasingly unavoidable
as technology and medicine continue to develop. Unfortunately, the computational cost of accurate
electronic structure calculations grows extremely fast with the number of particles, greatly limiting
the use of the state\hyp{}of\hyp{}the\hyp{}art first\hyp{}principles theories 
when large and complex chemical systems such as
solutions and biomolecules come into play.\cite{Tomasi.Mennucci.Cammi.ChemRev.2005}
%Although the advent in the computer technology 
%brought some hopes, it is still insufficient to follow up the expensive needs
%of computational chemistry.
Due to this reason, it has been of particular importance nowadays 
to develop bridging theoretical techniques that link
extremely accurate but localized pieces of information 
with certain knowledge of an atomistic organisation and intermolecular interaction
on a relatively large spatial scales, typically ranging beyond few nanometers.

The development of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM)
technique was a revolution in computational chemistry and allowed insightful theoretical studies
of quantum mechanical effects in biological systems because the high\hyp{}level
and costly quantum mechanical calculations
could be combined with low\hyp{}level and cheap classical dynamics.\cite{Warshel.Levitt.JMolBiol.1976,Senn.Thiel.Angew.2009}
%That was possible due to the fruitful incorporation of accurate \emph{ab initio} methods and a 
%semi\hyp{}classical notion of molecular motion combined with extensive experimental and semi\hyp{}empirical
%refinings.
Understanding physics that rules intermolecular interactions\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
helped to develop
a wide family of force fields, and served as a basis for embedding Hamiltonian
within an effective environment and fragmentation methods.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}
%The size of quantum level can be extended without compromising accuracy and maintaining good performance
%by using vast family of fragmentation methods, such as effective fragment potentials (EFP).
%The most difficult to tackle part of Hamiltonian can also be embedded within a widely understood effective environment
%leading to various embedding methodologies, realising similar goals.

However, even with the modern advances in computational chemistry, 
%accurate description of environmental effects
accurate description of phenomena due to complex heterogeneous molecular environments 
remains a challenge in the following areas: (i) describing the electronically excited states 
(electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,
Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017} 
Interface between QM and MM levels is usually realized by electrostatic embedding
and only rarely includes the Pauli exclusion principle and dispersion.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
because of complicated nature of intermolecular interactions, that theoretically
cannot be solely envisioned as a sum of electrostatic, dispersion and Pauli effects.
Perhaps the most frequently used efficient and reasonably accurate \emph{ab initio} fragmentation methods,
the effective fragment potential (EFP) methods, are limited to the ground\hyp{}state
chemistry at the Hartree\hyp{}Fock (HF)\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Nguyen.Pachter.Day.JCP.2014}.
Unfortunately, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.
%Therefore, Hamiltonian embedding is often realised by Coulombic embedding with polarization
%of QM level, and only rarely the Pauli exclusion principle is taken into account. Although
%such approach is very accurate in certain cases, it is not guaranteed that it is sufficient
%for all cases.

Most of efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{} (or force field\hyp{})based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wide scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,
though still not solved exactly.\cite{Holas.March.PhysRevA.1991}
%that seem to have an important advantage over the standard energetic approach.
With connection to the above, it would be incredibly insightful 
to establish a first\hyp{}principles connection bewteen the one\hyp{}particle
density and the perturbation in the environment through introducing the
\emph{density matrix susceptibility} (DMS) tensors. % because knowing the density distribution
%is equivalent to knowing the state of the molecule.
The main postulate of the present work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%and induced dipole moments $\delta {\BM\upmu}_i$ due to electric fields ${\bf E}_i$ 
%at the centroids of the occupied molecular orbitals,
%
\begin{equation*} \label{e:initial-model}
% \delta {\bf D} \approx \sum_i^{\text{Sites}} \text{[DMS of the $i$th site]} 
  \delta {\bf D} \approx \sum_i^{\text{Sites}} {\bf B}_i
  \cdot  \delta {\bf P}_i \;.
  %\rightarrow \sum_i^{\rm occ} {\bf B}^{(i)}_{\mu\nu} \cdot {\bf E}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential, in principle applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
becuase the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.
%
%a simple first\hyp{}principles model for prediction 
%of the perturbation\hyp{}induced one\hyp{}electron density
%change of structurally unrelaxed molecule is proposed here.

@@

It seems intuitive that $\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
%Unfortunately, such a mathematical problem is very complicated due to its high\hyp{}order tensor nature
%with respect to the unknown $\delta {\bf D}$ and no approximate analytical solutions have been proposed
%up to date. Therefore, in this paper we adopt a somewhat different
%approach that will require \emph{a priori} knowledge from training set of known cases.
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, %which is the first non\hyp{}vanishing moment,
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} \cdot {\mathbb{M}({\bf r}_Q)}
         \right] } \;.
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is AO\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = -e\tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this work we shall use the special open bold font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent on the origin.

In the Hartree\hyp{}Fock theory,\cite{Roothaan.RevModPhys.1951} ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C} \cdot {\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right] \cdot
                                   \left[ {\bf 1} + {\bf v}^\dagger \cdot {\bf v} \right]^{-1} \cdot
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)} + 
                        {\bf D}^{(0)} \cdot {\BM\Delta}^\dagger \cdot \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C} \cdot {\bf C}^{(0)\dagger} + {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \nonumber \\
           &\qquad - {\bf D}^{(0)} \cdot \delta {\bf C} \cdot {\bf C}^{{(0)}\dagger} 
                   - {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \cdot {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger} \cdot {\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated otherwise.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%%
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
%\end{equation}
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger} \cdot {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This guarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that
%%
%\begin{subequations}
% \begin{align}
%  \Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  
%    \right] }  &
%%
%  \nonumber\\ 
%  +\Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger}
%    \right] }
% &=
%2 \Tr{ 
%    \left[ 
%         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
%   \right] }  \;, 
%%
%\\
%%
%  \Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}
%    \right] }  &
%%
%   \nonumber\\
% +\Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger} \cdot {\bf D}^{(0)}
%    \right] }
% &= 
%2 \Tr{ 
%    \left[ 
%         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
%   \right] } \;.
%%
% \end{align}
%\end{subequations}
%%

>>>>

In this work, the electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibilities
interacting with electric fields. The analytical approximate expressions were derived
at the Hartree\hyp{}Fock theory, which served as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron density of water molecule in spatially non\hyp{}uniform electric field distribution
generetated by point charges. The electric field\hyp{}induced interaction energies,
dipole moments and quadrupole moments
are quantitatively described by the present theory in weak, moderate and even very strong electric fields (0.003 -- 0.07 a.u.).
It is believed that the proposed model could open
new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems
at arbitrary level of theory.
%at post\hyp{}Hartree\hyp{}Fock levels of theory.
\end{abstract}

@@

Molecular\hyp{}level exploration of condensed phase phenomena becomes increasingly unavoidable
as technology and medicine continue to develop. Unfortunately, the computational cost of accurate
electronic structure calculations grows extremely fast with the number of particles, greatly limiting
the use of the state\hyp{}of\hyp{}the\hyp{}art first\hyp{}principles theories 
when large and complex chemical systems such as
solutions and biomolecules come into play.\cite{Tomasi.Mennucci.Cammi.ChemRev.2005}
Due to this reason, it has been of particular importance nowadays 
to develop bridging theoretical techniques that link
extremely accurate but localized pieces of information 
with certain knowledge of an atomistic organisation and intermolecular interaction
on a relatively large spatial scales, typically ranging beyond few nanometers.

The development of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM)
technique was a revolution in computational chemistry and allowed insightful theoretical studies
of quantum mechanical effects in biological systems because the high\hyp{}level
and costly quantum mechanical calculations
could be combined with low\hyp{}level and cheap classical dynamics.\cite{Warshel.Levitt.JMolBiol.1976,Senn.Thiel.Angew.2009}
Understanding physics that rules intermolecular interactions\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
helped to develop
a wide family of force fields, and served as a basis for embedding Hamiltonian
within an effective environment and fragmentation methods.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}

However, even with the modern advances in computational chemistry, 
accurate description of phenomena due to complex heterogeneous molecular environments 
remains a challenge in the following areas: (i) describing the electronically excited states 
(electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,
Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017} 
Interface between QM and MM levels is usually realized by electrostatic embedding
and only rarely includes the Pauli exclusion principle and dispersion.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
because of complicated nature of intermolecular interactions, that theoretically
cannot be solely envisioned as a sum of electrostatic, dispersion and Pauli effects.
Perhaps the most frequently used efficient and reasonably accurate \emph{ab initio} fragmentation methods,
the effective fragment potential (EFP) methods, are limited to the ground\hyp{}state
chemistry at the Hartree\hyp{}Fock (HF)\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Nguyen.Pachter.Day.JCP.2014}.
Unfortunately, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.

Most of efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{} (or force field\hyp{})based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wide scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,
though still not solved exactly.\cite{Holas.March.PhysRevA.1991}
With connection to the above, it would be incredibly insightful 
to establish a first\hyp{}principles connection bewteen the one\hyp{}particle
density and the perturbation in the environment through introducing the
\emph{density matrix susceptibility} (DMS) tensors. 
The main postulate of the present work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%
\begin{equation*} \label{e:initial-model}
  \delta {\bf D} \approx \sum_i^{\text{Sites}} {\bf B}_i
  \cdot  \delta {\bf P}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential, in principle applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
becuase the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.

@@

It seems intuitive that $\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, 
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} \cdot {\mathbb{M}({\bf r}_Q)}
         \right] } \;.
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is AO\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = -e\tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this work we shall use the special open bold font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent on the origin.

In the Hartree\hyp{}Fock theory,\cite{Roothaan.RevModPhys.1951} ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C} \cdot {\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right] \cdot
                                   \left[ {\bf 1} + {\bf v}^\dagger \cdot {\bf v} \right]^{-1} \cdot
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)} + 
                        {\bf D}^{(0)} \cdot {\BM\Delta}^\dagger \cdot \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C} \cdot {\bf C}^{(0)\dagger} + {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \nonumber \\
           &\qquad - {\bf D}^{(0)} \cdot \delta {\bf C} \cdot {\bf C}^{{(0)}\dagger} 
                   - {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \cdot {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger} \cdot {\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated otherwise.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger} \cdot {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This guarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that
%
\begin{equation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
   \right] }  
-
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
   \right] } \;.
\end{equation}
%
Here we introduced the auxiliary tensors
%
\begin{subequations}
 \begin{align}
   \widetilde{\mathbb{M}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}} \cdot {\bf D}^{(0)} \;, \\
   \widetilde{\mathbb{L}}  &\equiv \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \;.
 \end{align}
\end{subequations}
%
to simplify the notation. Hence, we have
%
\begin{equation} \label{e:dmu-l-vector}
  \frac{1}{4} 
 \delta {\BM{\upmu}} %({\bf r}_0) 
   =
   \Tr{ 
    \left[ 
         \widetilde{\mathbb{L}} \cdot \delta {\bf C}
    \right] }
   %
   \equiv \sum_{i}^{\rm occ} \sum_\alpha {\bf l}_{i\alpha} \delta C_{\alpha i} \;,
\end{equation}
%
where the Cartesian vector
%
\begin{equation}\label{e:l-vector}
 {\bf l}_{i\alpha} \equiv \left[ \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \right]_{i\alpha} 
      = \left[  {\bf C}^\dagger \cdot \mathbb{M} \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \right]_{i\alpha}
\end{equation}
%
and $i$ runs over occupied unperturbed MO's.
Note that the intermediate quantity $\widetilde{\mathbb{L}}$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$
due to the presence of trace operation, unless the electric field is uniform.
Let us partition the induced dipole moment into separate contributions
from the occupied MO's, i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i
\end{equation}
%
where
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i = {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed polarizability tensor
associated with the $i$th MO whereas ${\bf F}({\bf r}_i)$ is the electric field evaluated at the $i$th MO centroid of charge
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
We can now insert the above result into Eq.~\eqref{e:dmu-l-vector} 
and after recasting the right hand side in a matrix form
we get
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^T \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i$ is a matrix of size $(n \times 3)$ and $\delta{\bf c}_i$ is just the $i$th column of the
change in the LCAO-MO matrix due to the polarization process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^T = \frac{1}{4}
            {\bf F}({\bf r}_i)  \cdot {\BM{\alpha}}_i \cdot 
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix 
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^T \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^T 
\end{equation}
%
and has size $(3\times n)$. Note that the \emph{right inverse} of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^T$ is singular due to the fact that usually
$n > 3$.

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha}^{(i;1)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha;w}^{(i;1)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(1)$' underlines the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i;1)}_{\alpha\beta} = 
                               C_{\alpha i} {\bf b}_{\beta}^{(i;1)} + C_{\beta i} {\bf b}_\alpha^{(i;1)} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma}^{(i;1)}
 \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i;1)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with the $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i;1)}$
are unique properties of electronically unperturbed molecule.

Eq.~\eqref{e:final-model.HF}, which is the final result of the previous section, 
can be used as a guideline to construct a general model of density matrix polarization.
Quadratic terms with respect to the electric field as well as linear terms with respect to
electric field gradient can be added to the formulation, which leads to
%
\begin{multline}\label{e:final-model.General}
 \delta D_{\alpha\beta} = \sum_{i }^M \Big\{
                                      {\bf B}^{(i ; 1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \\
%                       + \sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)
                        +             {\bf B}^{(i ; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) 
                        + \ldots \Big\}
\end{multline}
%
In the above equation, the density matrix susceptibilities
${\bf B}^{(i; R)}_{\alpha\beta}$, distributed over $M$ sites,
denote the $R$-th order response with respect to the electric field.
Note that
${\bf B}^{(i; 1)}_{\alpha\beta}$ is a vector of length $3$ whereas ${\bf B}^{(i; 2)}_{\alpha\beta}$
is $(3\times 3)$ matrix. Also, the distributed expansion centres
(labeled as $i$) are generalized to any point in the molecule.\footnote{More general
contribution due to square of electric field could be $\sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)$. However, for simplicity we neglect the contributions from different sites $i$ and $j$
in the present discussion and leave it for future development.
}
